Supplementary Material for the article \textit{Quotient Normalized Maximum Likelihood Criterion for Learning Bayesian Network Structures}

### 2017-11-23 AnswerToReviewers.txt

We thank reviewers for their feedback. While there appears to be a big spread in overall ratings, we are happy to see that the reviews are unanimously positive about the theoretical properties of the proposed model selection criterion, qNML, that is decomposable (unlike NML), consistent, score equivalent (unlike fNML), parameter free (unlike BDeu or BDq), regular (unlike BDeu) and possibly a reasonable approximation to the NML criterion.

Our original motivation for looking for a new model selection criterion was, however, extremely practical. Most of our "clients" want to use Bayesian networks due to their intuitive interpretability, and for the insight Bayesian networks provide to the domain. Unfortunately, that is where existing methods too often appear to fail us. To practitioner's disappointment, BIC often yields an empty network that is hard to sell, say, to a social scientist that has spent months to painstakingly collect his/her (often not very big) data set. Unfortunately, our experiments using Structural Hamming Distance (SHD) does not fully reveal this situation, since the empty network can on average still be relatively competitive in terms of SHD. 
Naturally, providing very "false insight" is also undesirable so the models should also be reasonably good in prediction. To our experience, the  qNML is closest to strike a good balance for providing insight while still being predictively well performing, the fact that for example reviewer 1 appears to have appreciated. While probably chosen for communicative clarity, we feel that 
"it does not work in practice" by reviewer 2 is not a best possible summary of the practical value of qNML, which unlike other methods(!), never performs very poorly.

Clearly, none of the criteria perform the best for all sample sizes and purposes (prediction vs. SHD), which might even suggest picking different criteria for different situations and needs. We like, like reviewer 3, feel that in this situation, proposing a theoretically well behaved and empirically validated criterion is indeed an interesting contribution to the field.

Reviewer 1 gave some detailed comments that are easy to take into account in the final version of the paper. In the longer version of the paper, the comparison with the BDq, as suggested by reviewer 3, is a natural thing to do. The issue with BDq is that it features a hyper-parameter and our preliminary results indicate that BDq can be very sensitive to this parameter. Therefore, for more meaningful comparison, the not yet thoroughly studied parameter sensitivity issue must also be addressed. For including it to this conference paper might be to much of a squeeze.

### 2017-11-27 AnswerToReviewers.txt

We thank reviewers for their feedback. While there appears to be a big spread in overall ratings, we are happy to see that the reviews are unanimously positive about the theoretical properties of the proposed model selection criterion, qNML, that is decomposable (unlike NML), consistent, score equivalent (unlike fNML), parameter free (unlike BDeu or BDq), regular (unlike BDeu) and possibly a reasonable approximation to the NML criterion.

Our original motivation for looking for a new model selection criterion was, however, extremely practical. Most of our "clients" want to use Bayesian networks due to their intuitive interpretability, and for the insight Bayesian networks provide to the domain. Unfortunately, that is where existing methods too often appear to fail us. To practitioner's disappointment, BIC often yields an empty network that is hard to sell, say, to a social scientist that has spent months to painstakingly collect his/her (often not very big) data set. Unfortunately, our experiments using Structural Hamming Distance (SHD) does not fully reveal this situation, since the empty network can on average still be relatively competitive in terms of SHD. 
Naturally, providing very "false insight" is also undesirable so the models should also be reasonably good in prediction. To our experience, the  qNML is closest to strike a good balance for providing insight while still being predictively well performing, the fact that for example reviewer 1 appears to have appreciated. While probably chosen for communicative clarity, we feel that 
"it does not work in practice" by reviewer 2 is not a best possible summary of the practical value of qNML, since it appears as a best method in SHD (see Figure 3) from medium sample sizes up. Furthermore, unlike BIC and BDeu, it seldom performs very poorly, and, like pointed out by the reviewer 1, it has both theoretical and practical advantages over the other robust criterion, fNML.

Clearly, none of the criteria perform the best for all sample sizes and purposes (prediction vs. SHD), which might even suggest picking different criteria for different situations and needs. We like, like reviewer 3, feel that in this situation, proposing a theoretically well behaved and empirically validated criterion is indeed an interesting contribution to the field - a contribution that the Bayesian network learning community should know about.

### 2017-11-27 AnswerToReviewers.txt

We thank reviewers for their feedback. While there appears to be a big spread in overall ratings, we are happy to see that the reviews are unanimously positive about the theoretical properties of the proposed model selection criterion, qNML, that is decomposable (unlike NML), consistent, score equivalent (unlike fNML), parameter free (unlike BDeu or BDq), regular (unlike BDeu) and possibly a reasonable approximation to the NML criterion.

Reviewer 1 gave some detailed comments that are easy to take into account in the final version of the paper. In the longer version of the paper, the comparison with the BDq, as suggested by reviewer 3, is a natural thing to do. The issue with BDq is that it features a hyper-parameter and our preliminary results indicate that BDq can be very sensitive to this parameter. Therefore, for more meaningful comparison, the not yet thoroughly studied parameter sensitivity issue must also be addressed. For including it to this conference paper might be too much of a squeeze.

Our original motivation for looking for a new model selection criterion was extremely practical. Most of our "clients" want to use Bayesian networks due to their intuitive interpretability, and for the insight Bayesian networks provide to the domain. Unfortunately, that is where existing methods too often appear to fail us. To practitioner's disappointment, BIC often yields an empty network that is hard to sell, say, to a social scientist that has spent months to painstakingly collect his/her (often not very big) data set. Unfortunately, our experiments using Structural Hamming Distance (SHD) does not fully reveal this situation, since the empty network can on average still be relatively competitive in terms of SHD.

Naturally, providing very "false insight" is also undesirable so the models should also be reasonably good in prediction. To our experience, the  qNML is closest to strike a good balance for providing insight while still being predictively well performing, the fact that for example reviewer 1 appears to have appreciated. While probably chosen for communicative clarity, we feel that 
"it does not work in practice" by reviewer 2 is not a best possible summary of the practical value of qNML, since it appears as a best method in SHD (see Figure 3) from medium sample sizes up. Furthermore, taking a closer look at the empirical results we can see the following:

- based on the average rank, qNML is better in prediction than BIC which produces the worst predictions on large sample size settings.
- qNML obtains always a better rank than fNML in structure learning experiments (see Figure 3). When averaged over all the sample sizes, fNML obtains the best rank in prediction with qNML being the second to best. However, the standard deviation of fNML's rank is 0.76 which is around 77% larger than qNML's (0.43). Also, as Table 3 suggests, qNML obtains this competitive performance in prediction usually with considerably simpler models when compared to fNML.   
- BDeu is bested by all the other methods nearly always in prediction. In structure learning, there are couple settings with moderate sample sizes where BDeu has a better average rank than qNML but otherwise qNML outperforms BDeu.

Clearly, none of the criteria perform the best for all sample sizes and purposes (prediction vs. SHD), which might even suggest picking different criteria for different situations and needs. We, like reviewer 3, feel that in this situation, proposing a theoretically well behaved and empirically validated criterion is indeed an interesting contribution to the field - a contribution that the Bayesian network learning community should know about.

### 2017-11-28 AnswerToReviewers.txt

We thank reviewers for their feedback. While there appears to be some spread in overall ratings, we are happy to see that the reviews are unanimously positive about the theoretical properties of the proposed model selection criterion, qNML, that is decomposable (unlike NML), consistent, score equivalent (unlike fNML), parameter free (unlike BDeu or BDq), regular (unlike BDeu) and possibly a reasonable approximation to the NML criterion.

Reviewer 1 gave some detailed comments that are easy to take into account in the final version of the paper. In a journal version of the paper, the comparison with the BDq, as suggested by reviewer 3, would be a natural thing to do. The issue with BDq is that it features a hyper-parameter and our preliminary results indicate that like BDeu, BDq too can be extremely sensitive to the choice of the hyperparameter. Therefore, for more meaningful comparison, the not yet thoroughly studied parameter sensitivity issue with BDq should also be addressed.

About the practicality of the proposed method: Our original motivation for looking for a new model selection criterion was actually extremely practical. We have quite a lot of experience of working with users (researchers in applied fields), and they want to use Bayesian networks due to the insight Bayesian networks provide into their domain. Unfortunately, that is where the existing methods too often appear to fail. For example, to the practitioners' disappointment, BIC often yields an empty network. This is not useful at all, say, to a social scientist that has spent months to painstakingly collect his/her (often not very big) data set. Our experiments using Structural Hamming Distance (SHD) do not fully reveal this situation, since the empty network can actually be relatively competitive in terms of SHD.

Obviously, providing more but misleading "insight" is even worse. So the models should also capture the underlying distribution, which we suggest to measure by prediction performance. What we are particularly happy about is that qNML finds a good balance by providing insight while still performing well in predictive tasks -- as pointed out by reviewer 1.

Based on these points, "it does not work in practice" (reviewer 2) is perhaps not really an accurate summary of the practicality of qNML. Besides the above points about balanced performance, it is the best method in SHD (see Figure 3) from medium sample sizes up.

Further remarks about the empirical results:

- based on the average rank, qNML is better in prediction than BIC which produces the worst predictions on large sample size settings (Table 2)
- qNML obtains always a better rank than fNML in structure learning experiments (Fig. 3).
- when averaged over all the sample sizes, fNML obtains the best rank in prediction with qNML being the second to best. However, the standard deviation of fNML's rank is 0.76 which is larger than qNML's (0.43) (data from Table 2; averages and variances not shown explicitly in the paper)
- qNML obtains this competitive performance in prediction with considerably simpler models when compared to fNML (Table 3).
- BDeu is outperformed by all the other methods in almost all prediction experiments (Table 2).
- In structure learning, qNML outperforms BDeu except in a few settings with moderate sample sizes (Fig. 3).

Clearly, none of the criteria perform the best for all sample sizes and purposes (prediction vs. SHD), which might even suggest picking different criteria for different situations and needs. We, like reviewer 3, feel that in this situation, proposing a theoretically well behaved and empirically robust criterion is indeed an interesting contribution to the field - a contribution that the Bayesian network learning community is likely to be interested to know about.

### 2018-02-12 bn.tex
ref to right appendix
address minor comment

Bayesian learning of Bayesian network structures is based on the
posterior probability $P(G|D,\alpha)$, where $\alpha$ denotes the
hyperparameters for the model parameters $\theta$, and the $D$ is a
collection of $N$ $n$\nobreakdash-dimensional i.i.d. data vectors
collected to a $N\times n$ design matrix. We use the notation $D_i$ to
denote the $i^\text{th}$ column of the data matrix and notation $D_V$ to denote
the columns that correspond to the variable subset $V$. We also write
$D_{i,G_i}$ for $D_{\{i\}\cup G_i}$ and denote the entries of the
column $i$ on the rows on which the parents $G_i$ contain the value
configuration number by $D_{i,G_i=j}$, $j\in\{1,\ldots,q_i\}$.

@@

Recently, Suzuki~\cite{Suzuki2017} discussed the theoretical
properties of the BDeu score and showed that in certain settings BDeu
has tendency to add more and more parent variables for a child node
even though the empirical conditional entropy of the child given the
parents has already reached zero. In more detail, assume that in our
data $D$, the values of $X_i$ are completely determined by variables
in set $Z$, so that the empirical entropy $H_N(X_i | Z) = 0$. Now, if we can further
find one or more variables, denoted by $Y$, whose values are
determined completely by the variables in $Z$, then BDeu will prefer
the set $Z\cup Y$ over $Z$ alone as the parents of $X_i$. Suzuki
argues that this kind of behaviour violates \textit{regularity} in model
selection as the more complex model is preferred over a simpler one
even though it does not fit the data any better. The phenomenon seems
to stem from the way the hyperparameters for the Dirichlet
distribution are chosen in BDeu as using Jeffreys' prior,
$\theta_{ijk}\sim Dir(\frac{1}{2},\ldots,\frac{1}{2})$, does not suffer
from this anomaly. However, using Jeffreys' prior causes marginal
likelihood score not to be score equivalent. In Section \ref{sec:regularity}, we will give the formal definition of regularity and state that qNML is regular. In addition, we provide a proof of regularity for fNML criterion, which has not appeared in the literature before. The detailed proofs can be found in Appendix A in the Supplementary Material.

@@

Consequently, Silander et al.~\cite{cosco.pgm08a}
suggested solving the BDeu parameter sensitivity problem by using the
NML-code for the column partitions, i.e., changing the Bayesian mixture
in equation~(\ref{eqn:bayesmix}) to
\begin{equation}
P^1_{NML}(D_{i,G_i=j};G)=\frac{P(D|\hat\theta(D_{i,G_i=j};G))}{\sum_{D'}{P(D';\hat\theta(D';G))}},
\end{equation}
where $D'\in{\{1,\ldots,r_i\}}^{|D_{i,G_i=j}|}$.  The logarithm of the
denominator is often called a regret, since it indicates the extra
code length needed compared to the code length obtained using the (a
priori unknown) maximum likelihood parameters. The regret for
$P^1_{NML}$ depends only on the length $N$ of the categorical data
vector with $r$ different categorical values,
\begin{equation}
reg(N,r)=\log \sum_{D\in \{1,\ldots,r\}^N} P(D|\hat\theta(D)).
\end{equation}
While the naive
computation of the regret is still prohibitive, Silander et al.\
approximate it efficiently using a so-called Szpankowski
approximation~\cite{cosco.aistat03}:
\begin{eqnarray}
\label{eqn:szp1}
\lefteqn{reg(N,r) \approx \frac{\sqrt{2} r \Gamma{\left(\frac{r}{2} \right)}}
                               {3 \sqrt{N} \Gamma{\left(\frac{r-1}{2}  \right)}}} \\
&&+ \left(\frac{r-1}{2}\right) \log{\left (\frac{N}{2} \right )}
- \log \Gamma{\left(\frac{r}{2} \right)} + \frac{1}{2} \log{\left (\pi \right )}\nonumber\\
&&- \frac{r^{2} \Gamma^{2}{\left(\frac{r }{2} \right)}}
         {9N \Gamma^{2}{\left(\frac{r-1}{2}\right)}}
+ \frac{2r^3-3r^2-2r+3}{36N}\nonumber.
\end{eqnarray}

>>>>

Score-based Bayesian learning of Bayesian network structures evaluates the goodness of different structures $G$ using their
posterior probability $P(G|D,\alpha)$, where $\alpha$ denotes the
hyperparameters for the model parameters $\theta$, and the $D$ is a
collection of $N$ $n$\nobreakdash-dimensional i.i.d. data vectors
collected to a $N\times n$ design matrix. We use the notation $D_i$ to
denote the $i^\text{th}$ column of the data matrix and notation $D_V$ to denote
the columns that correspond to the variable subset $V$. We also write
$D_{i,G_i}$ for $D_{\{i\}\cup G_i}$ and denote the entries of the
column $i$ on the rows on which the parents $G_i$ contain the value
configuration number by $D_{i,G_i=j}$, $j\in\{1,\ldots,q_i\}$.

@@

Recently, Suzuki~\cite{Suzuki2017} discussed the theoretical
properties of the BDeu score and showed that in certain settings BDeu
has tendency to add more and more parent variables for a child node
even though the empirical conditional entropy of the child given the
parents has already reached zero. In more detail, assume that in our
data $D$, the values of $X_i$ are completely determined by variables
in set $Z$, so that the empirical entropy $H_N(X_i | Z) = 0$. Now, if we can further
find one or more variables, denoted by $Y$, whose values are
determined completely by the variables in $Z$, then BDeu will prefer
the set $Z\cup Y$ over $Z$ alone as the parents of $X_i$. Suzuki
argues that this kind of behaviour violates \textit{regularity} in model
selection as the more complex model is preferred over a simpler one
even though it does not fit the data any better. The phenomenon seems
to stem from the way the hyperparameters for the Dirichlet
distribution are chosen in BDeu as using Jeffreys' prior,
$\theta_{ijk}\sim Dir(\frac{1}{2},\ldots,\frac{1}{2})$, does not suffer
from this anomaly. However, using Jeffreys' prior causes marginal
likelihood score not to be score equivalent. In Section \ref{sec:regularity}, we will give the formal definition of regularity and state that qNML is regular. In addition, we provide a proof of regularity for fNML criterion, which has not appeared in the literature before. The detailed proofs can be found in Appendix B in the Supplementary Material.

@@

Consequently, Silander et al.~\cite{cosco.pgm08a}
suggested solving the BDeu parameter sensitivity problem by using the
NML-code for the column partitions, i.e., changing the Bayesian mixture
in equation~(\ref{eqn:bayesmix}) to
\begin{equation}
P^1_{NML}(D_{i,G_i=j};G)=\frac{P(D|\hat\theta(D_{i,G_i=j};G))}{\sum_{D'}{P(D'|\hat\theta(D';G))}},
\end{equation}
where $D'\in{\{1,\ldots,r_i\}}^{|D_{i,G_i=j}|}$.  The logarithm of the
denominator is often called a regret, since it indicates the extra
code length needed compared to the code length obtained using the (a
priori unknown) maximum likelihood parameters. The regret for
$P^1_{NML}$ depends only on the length $N$ of the categorical data
vector with $r$ different categorical values,
\begin{equation}
reg(N,r)=\log \sum_{D\in \{1,\ldots,r\}^N} P(D|\hat\theta(D)).
\end{equation}
While the naive
computation of the regret is still prohibitive, Silander et al.\
approximate it efficiently using a so-called Szpankowski
approximation~\cite{cosco.aistat03}:
\begin{eqnarray}
\label{eqn:szp1}
\lefteqn{reg(N,r) \approx \frac{\sqrt{2} r \Gamma{\left(\frac{r}{2} \right)}}
                               {3 \sqrt{N} \Gamma{\left(\frac{r-1}{2}  \right)}}} \\
&&+ \left(\frac{r-1}{2}\right) \log{\left (\frac{N}{2} \right )}
- \log \Gamma{\left(\frac{r}{2} \right)} + \frac{1}{2} \log{\left (\pi \right )}\nonumber\\
&&- \frac{r^{2} \Gamma^{2}{\left(\frac{r }{2} \right)}}
         {9N \Gamma^{2}{\left(\frac{r-1}{2}\right)}}
+ \frac{2r^3-3r^2-2r+3}{36N}\nonumber.
\end{eqnarray}
